### 2020-06-11 chapters/sota.tex
Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm

A time-series is a collection of data points indexed by time. Accordingly, the authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}. Regarding the concept of \textit{outlier}, the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. Hence, outlier detection refers to the process of finding anomalies in data. An anomaly relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

@@

\iffalse
ExtraN\\
point outlier\\
density-based outliers\\
predictability/forecast property based\\

HTM\\
point outlier, model-based, prediction\\
black-box ML algorithm\\
explain properties\\
use what authors say in their review\\

DeepAnT\\
point outlier, model-based, prediction\\
black box DL algorithm\\
explain properties\\
use what authors say in their review\\
\fi

@@

>>>>

A time-series is a collection of data points indexed by time. Accordingly, the authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}. Regarding the concept of \textit{outlier}, the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. In summary, outlier detection refers to the process of finding anomalies in data. An anomaly (or outlier) relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

@@

\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm}

Yang et. al propose a method for incremental detection of density-based point outliers \cite{Yang-Neighbor-Based-Pattern-Detection}, specifically for sliding windows scenarios. The authors propose Extra-N, an algorithm that detects point outliers based on the number of neighbors it has. The authors make use of the \textit{“predictability"} property of the expiration of existing objects. In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to. This technique allows the efficient maintenance of neighborship counts by pre-handling the eventual expiring effect of new data points.

Ideally we want to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has rather than the actual list of neighbors. The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors. The problem is that when a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one but we are unable to reach \textit{p\textsubscript{i}}'s neighbors without a list of neighbor pointers. The solution to this problem is exploiting the previously presented \textit{predictability} property. The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbors. The authors introduce the concept of \textit{"lifetime neighbor counts"} or \textit{lt\_cnt}. The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where he exists. More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s neighbors that are known to survive in the corresponding future window. The authors provide the following example: consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors, \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}. Applying the \textit{predictability} property makes it possible to compute the lifespan of all points. Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}. This implies that at \textit{W\textsubscript{i}}, 
\[ \textit{p\textsubscript{i}.lt\_cnt} = \{\textit{W\textsubscript{i}}: 3, \textit{W\textsubscript{i+1}}: 2, \textit{W\textsubscript{i+2}}: 0\} \]
The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point \textit{p\textsubscript{i}} check if the first or current element of \textit{p\textsubscript{i}.lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether a data point is an outlier or not.

\subsubsection*{Time complexity analysis}

To find outliers, Extra-N has to do a linear search through the list of points and check the neighbor counts. Hence, for outlier detection, this method shows \textit{O(n)} complexity with \textit{n} as the number of points. Additionally, for each new data point \textit{p\textsubscript{i}}, Extra-N has to perform a range query search centered on it to find its neighbors and increment their lifetime neighbor counts by one. The run-time complexity of a range query search is \textit{O(log(n))} \cite{range-query-log} with \textit{n} as the number of data points.

In summary, Extra-N shows logarithmic time complexity when updating the sliding window and linear time complexity when looking for outliers, both relative to the window size.

\subsubsection*{Space complexity analysis}
Extra-N has to keep a lifetime neighbor count dictionary-like object for each window data point. For that reason, we claim that Extra-N achieves linear memory consumption relative to the number of data points in the window.

\subsubsection*{Applicability to our Hypothesis}
Extra-N memory consumption grows linearly in regards to window size, so it does not fit our desired lightweight system architecture. Furthermore, updating the sliding window should be done in constant time, which is not the case for Extra-N. Hence, this algorithm will be of no use to us.

\subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
HTM\\
point outlier, model-based, prediction\\
black-box ML algorithm\\
explain properties\\
use what authors say in their review\\

\subsection{DeepAnT}
DeepAnT\\
point outlier, model-based, prediction\\
black box DL algorithm\\
explain properties\\
use what authors say in their review\\

@@

\subsection{A two window paradigm algorithm for change detection}
